{\bf {\Large Four decades of water quality change in the upper San Francisco Estuary}}

### 2016-10-19 sftrends_manu.Rnw
more site desription

Nutrient loading in \ac{sfe} is comparable to other locations that exhibit symptomatic effects of cultural eutrophication, although productivity has historically not been affected by excess nutrients \citep{Jassby02,Kimmerer12}.  Orthoposphate (PO$_4^{3-}$) and \ac{din} enter the Bay primarily through riverine sources in the north and municipal \ac{wwtp} inputs in the densely-populated area immediately surrounding \ac{sfe}.  Annual nutrient export from the Delta region as been estimated as approximateily 30 thousand kg d$-1$ of total nitrogen \citep[varying with flow,][]{Novick15}, with 90\% of ammonium (NH$_4^{+}$) originating solely from the Sacramento Regional \ac{wwtp} \citep{Jassby08}.  Although nitrogen inputs are considerable, several

POTW (publicly owned treatment works), nutrient loading relative to other locations, recent trends

The delta region...

>>>>

Nutrient loading in \ac{sfe} is comparable to other large estuaries that exhibit symptomatic effects of cultural eutrophication \citep[e.g., Chesapeake Bay,][]{Kemp05}.  Orthophosphate (PO$_4^{3-}$) and \ac{din} enter the Bay primarily through riverine sources in the north and municipal \ac{wwtp} inputs in the densely-populated area immediately surrounding \ac{sfe}.  Annual nutrient export from the Delta region has been estimated as approximateily 30 thousand kg d$^{-1}$ of total nitrogen \citep[varying with flow,][]{Novick15}, with 90\% of ammonium (NH$_4^{+}$) originating solely from the Sacramento Regional \ac{wwtp} \citep{Jassby08}.  Although nitrogen inputs are considerable, primary production is relatively low and not nutrient-limited \citep{Jassby02,Kimmerer12}.  The resistance of \ac{sfe} to the negative effects of eutrophication has been attributed to the unique physical and biological characteristics of the Bay, including strong tidal mixing that limits stratification \citep{Cloern96,Thompson08} and limits on phytoplankton growth from high turbidity and filter-feeding by bivalve mollusks \citep{Thompson08,Crauder16}.  Recent water quality trends have suggested that resistence of the system to nutrient inputs is decreasing given documented increases in chlorophyll biomass, increased occurrence of hypoxic conditions, and increased abundance of phytoplankton species associated with harmful algal blooms.  These recent changes have been attributed to variation in global sea surface temperatures associated with climate change \citep{Cloern07}, biological invasions \citep{Cohen98}, and departures from the historical flow record \citep{Enright09,Cloern12}.

The Delta region is of particular interest for understanding historical patterns and potential trajectories of water quality in response to nutrient inputs into the Bay.  The Delta is a mosaic of linked channels or tracts that receive, process, and transport inflows from the Sacramento and San Joaquin rivers.  Quantitative descriptions of nutrient dynamics in the Delta are complicated by the numerous sources that contribute input loads.  Sources are not limited to the watershed, such that nutrient concentrations can also originate from local wastewater inputs and internal loading from sediment processes.  Seasonal and annual variation in the delivery of water inflows, including water exports directly from the Delta, have further complicated the interpretation of trends in nutrient concentrations. A comprehensive evaluation of nutrient dynamics using mass-balance models showed that

Accordingly, quantitative descriptions of nutrient dynamics in the Delta must explicitly account for the effects of flow changes to better understand variation both within the Delta and potential mechanisms of downstream tranport.

Two data sources that provide multi-decadal time series of nutrients and daily flow records were used to develop a quantitative description fo

### 2016-10-19 sftrends_manu.tex
more site desription

\begin{singlespace}
\title{{\bf {\Large Four decades of water quality change in the upper San Francisco Estuary}}}
\author{
  {\bf {\normalsize Marcus W. Beck$^1$, David Senn$^2$, Phil Bresnahan$^2$, Emily Novick$^2$, James D. Hagy III$^1$,}}
  \\{\bf {\normalsize Thomas Jabusch$^2$}}
  \\\\{\textit {\normalsize $^1$USEPA National Health and Environmental Effects Research Laboratory}}
  \\{\textit {\normalsize Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32561}}
	\\{\textit {\normalsize Phone: 850-934-2480, Fax: 850-934-2401}}
	\\{\textit {\normalsize Emails: \href{mailto:beck.marcus@epa.gov}{beck.marcus@epa.gov}, \href{mailto:hagy.jim@epa.gov}{hagy.jim@epa.gov}}}
  \\\\{\textit {\normalsize $^2$San Francisco Estuary Institute}}
	\\{\textit {\normalsize 4911 Central Avenue, Richmond, CA 94804}}
	\\{\textit {\normalsize Phone: 510-746-7334, Fax: 510-746-7300}}
	\\{\textit {\normalsize Emails: \href{mailto:davids@sfei.org}{davids@sfei.org}, \href{mailto:philb@sfei.org}{philb@sfei.org}, \href{mailto:emilyn@sfei.org}{emilyn@sfei.org}, \href{mailto:thomasj@sfei.org}{thomasj@sfei.org}}}
  \vspace{1in} 
  \\ Version Date:   Thu Oct 13 08:56:33 2016 -0500
	}
\date{}
\maketitle
\end{singlespace}
\clearpage

@@

Nutrient loading in \ac{sfe} is comparable to other locations that exhibit symptomatic effects of cultural eutrophication, although productivity has historically not been affected by excess nutrients \citep{Jassby02,Kimmerer12}.  Orthoposphate (PO$_4^{3-}$) and \ac{din} enter the Bay primarily through riverine sources in the north and municipal \ac{wwtp} inputs in the densely-populated area immediately surrounding \ac{sfe}.  Annual nutrient export from the Delta region as been estimated as approximateily 30 thousand kg d$-1$ of total nitrogen \citep[varying with flow,][]{Novick15}, with 90\% of ammonium (NH$_4^{+}$) originating solely from the Sacramento Regional \ac{wwtp} \citep{Jassby08}.  Although nitrogen inputs are considerable, several

POTW (publicly owned treatment works), nutrient loading relative to other locations, recent trends

The delta region...

>>>>

\begin{singlespace}
\title{{\bf {\Large Four decades of water quality change in the upper San Francisco Estuary}}}
\author{
  {\bf {\normalsize Marcus W. Beck$^1$, David Senn$^2$, Phil Bresnahan$^2$, Emily Novick$^2$, James D. Hagy III$^1$,}}
  \\{\bf {\normalsize Thomas Jabusch$^2$}}
  \\\\{\textit {\normalsize $^1$USEPA National Health and Environmental Effects Research Laboratory}}
  \\{\textit {\normalsize Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32561}}
	\\{\textit {\normalsize Phone: 850-934-2480, Fax: 850-934-2401}}
	\\{\textit {\normalsize Emails: \href{mailto:beck.marcus@epa.gov}{beck.marcus@epa.gov}, \href{mailto:hagy.jim@epa.gov}{hagy.jim@epa.gov}}}
  \\\\{\textit {\normalsize $^2$San Francisco Estuary Institute}}
	\\{\textit {\normalsize 4911 Central Avenue, Richmond, CA 94804}}
	\\{\textit {\normalsize Phone: 510-746-7334, Fax: 510-746-7300}}
	\\{\textit {\normalsize Emails: \href{mailto:davids@sfei.org}{davids@sfei.org}, \href{mailto:philb@sfei.org}{philb@sfei.org}, \href{mailto:emilyn@sfei.org}{emilyn@sfei.org}, \href{mailto:thomasj@sfei.org}{thomasj@sfei.org}}}
  \vspace{1in} 
  \\ Version Date:   Tue Oct 18 17:52:14 2016 -0500
	}
\date{}
\maketitle
\end{singlespace}
\clearpage

@@

Nutrient loading in \ac{sfe} is comparable to other large estuaries that exhibit symptomatic effects of cultural eutrophication \citep[e.g., Chesapeake Bay,][]{Kemp05}.  Orthophosphate (PO$_4^{3-}$) and \ac{din} enter the Bay primarily through riverine sources in the north and municipal \ac{wwtp} inputs in the densely-populated area immediately surrounding \ac{sfe}.  Annual nutrient export from the Delta region has been estimated as approximateily 30 thousand kg d$^{-1}$ of total nitrogen \citep[varying with flow,][]{Novick15}, with 90\% of ammonium (NH$_4^{+}$) originating solely from the Sacramento Regional \ac{wwtp} \citep{Jassby08}.  Although nitrogen inputs are considerable, primary production is relatively low and not nutrient-limited \citep{Jassby02,Kimmerer12}.  The resistance of \ac{sfe} to the negative effects of eutrophication has been attributed to the unique physical and biological characteristics of the Bay, including strong tidal mixing that limits stratification \citep{Cloern96,Thompson08} and limits on phytoplankton growth from high turbidity and filter-feeding by bivalve mollusks \citep{Thompson08,Crauder16}.  Recent water quality trends have suggested that resistence of the system to nutrient inputs is decreasing given documented increases in chlorophyll biomass, increased occurrence of hypoxic conditions, and increased abundance of phytoplankton species associated with harmful algal blooms.  These recent changes have been attributed to variation in global sea surface temperatures associated with climate change \citep{Cloern07}, biological invasions \citep{Cohen98}, and departures from the historical flow record \citep{Enright09,Cloern12}.

The Delta region is of particular interest for understanding historical patterns and potential trajectories of water quality response to nutrient inputs into the Bay.  The Delta is a mosaic of linked channels or tracts that receive, process, and transport inflows from the Sacramento and San Joaquin rivers to the lower Bay.  Quantitative descriptions of nutrient dynamics in the Delta are complicated by the numerous sources that contribute input loads.  Sources are not limited to the watershed, such that local inputs from wastewater and internal loading from sediment processes also represent important contributions.  Seasonal and annual variation in the delivery of water inflows, including water exports directly from the Delta, have complicated the interpretation of trends in nutrient concentrations. Accordingly, quantitative descriptions of nutrient dynamics in the Delta must explicitly account for the effects of flow changes to better understand variation both within the Delta and potential mechanisms of downstream tranport.

Two data sources that provide multi-decadal time series of nutrients and daily flow records were used to develop a quantitative description fo
